---
title: "Statistical Rethinking: Chapter 13: Adventures in Covariance"
output: html_notebook
---

# Varying slopes by construction

## Simulate the population

Define the population of cafes that the robot might visit. 

```{r}
a <- 3.5 # average morning wait time
b <- (-1) # average difference afternoon wait time
sigma_a <- 1 # std dev in intercepts
sigma_b <- 0.5 # std dev in slopes
rho <- (-0.7) # correlation between intercepts and slopes

Mu <- c(a, b)

# build the variance covariance matrix
cov_ab <- sigma_a * sigma_b * rho
Sigma <- matrix(c(sigma_a^2, cov_ab, cov_ab, sigma_b^2), ncol = 2)

## or
sigmas <- c(sigma_a, sigma_b) # standard deviation
Rho <- matrix(c(1, rho, rho, 1), nrow = 2) # correlation matrix

# now matrix multiply to get covariance matrix
Sigma <- diag(sigmas) %*% Rho %*% diag(sigmas)
```

Now we simulate cafes:

```{r}
N_cafes <- 20
library(MASS)
set.seed(5) # used to replicate example
vary_effects <- mvrnorm(N_cafes, Mu, Sigma)

a_cafe <- vary_effects[,1]
b_cafe <- vary_effects[,2]

plot( a_cafe , b_cafe , col=rangi2 ,
    xlab="intercepts (a_cafe)" , ylab="slopes (b_cafe)" )

# overlay population distribution
library(ellipse)
for ( l in c(0.1,0.3,0.5,0.8,0.99) )
    lines(ellipse(Sigma,centre=Mu,level=l),col=col.alpha("black",0.2))
```

Now fit the model:

```{r}
library(rethinking)

N_visits <- 10
afternoon <- rep(0:1,N_visits*N_cafes/2)
cafe_id <- rep( 1:N_cafes , each=N_visits )
mu <- a_cafe[cafe_id] + b_cafe[cafe_id]*afternoon
sigma <- 0.5  # std dev within cafes
wait <- rnorm( N_visits*N_cafes , mu , sigma )
d <- data.frame( cafe=cafe_id , afternoon=afternoon , wait=wait )

## R code 13.11
R <- rlkjcorr( 1e4 , K=2 , eta=2 )
dens( R[,1,2] , xlab="correlation" )

m13.1 <- map2stan(
  alist(wait ~ dnorm(mu, sigma),
        mu <- a_cafe[cafe] + b_cafe[cafe] * afternoon,
        c(a_cafe, b_cafe)[cafe] ~ dmvnorm2(c(a, b), sigma_cafe, Rho),
        a ~ dnorm(0, 10),
        b ~ dnorm(0, 10),
        sigma_cafe ~ dcauchy(0, 2),
        sigma ~ dcauchy(0, 2),
        Rho ~ dlkjcorr(2)),
  data = d,
  iter = 5000,
  warmup = 2000,
  chains = 2
)

post <- extract.samples(m13.1)
dens(post$Rho[ , 1, 2])
```

# Example: Admission decisions and gender

We want to model the admission decisions of graduate programs at UC Berkely but accounting for mean difference between departments as well as variation in how they treat male and female applicants.

```{r}
library(rethinking)
data("UCBadmit")
d <- UCBadmit
d$male <- ifelse(d$applicant.gender == "male", 1, 0)
d$dept_id <- coerce_index(d$dept)

m13.2 <- map2stan(
  alist(admit ~ dbinom(applications, p),
        logit(p) <- a_dept[dept_id] + bm * male,
        a_dept[dept_id] ~ dnorm(a, sigma_dept),
        a ~ dnorm(0, 10),
        bm ~ dnorm(0, 1),
        sigma_dept ~ dcauchy(0, 2)),
  data = d,
  warmup = 500,
  iter = 4500,
  chains = 3
)
precis(m13.2, depth = 2)
```

## Varying effects of being male

Now let's consider the gender bias among departments:

```{r}
m13.3 <- map2stan(
  alist(admit ~ dbinom(applications, p),
        logit(p) <- a_dept[dept_id] + bm_dept[dept_id] * male,
        c(a_dept, bm_dept)[dept_id] ~ dmvnorm2(c(a, bm), sigma_dept, Rho),
        a ~ dnorm(0, 10),
        bm ~ dnorm(0, 1),
        sigma_dept ~ dcauchy(0, 2),
        Rho ~ dlkjcorr(2)),
  data = d,
  warmup = 1000,
  iter = 5000,
  chains = 4,
  cores = 4
)
plot(m13.3)
precis(m13.3, depth = 2)
plot(precis(m13.3, pars = c("a_dept", "bm_dept"), depth = 2))
```

To compare models, we'll fit another model without gender.

```{r}
m13.4 <- map2stan(
  alist(admit ~ dbinom(applications, p),
        logit(p) <- a_dept[dept_id],
        a_dept[dept_id] ~ dnorm(a, sigma_dept),
        a ~ dnorm(0, 10),
        sigma_dept ~ dcauchy(0, 2)),
  data = d,
  warmup = 1000,
  iter = 5000,
  chains = 4,
  cores = 4
)

compare(m13.2, m13.3, m13.4)
```

# Example: Cross-classified chimpanzees with varying slopes

Let's fit the chimpanzee model with varying slopes for actors and blocks:

```{r}
library(rethinking)
data("chimpanzees")
d <- chimpanzees
d$recipient <- NULL
d$block_id <- d$block

m13.6 <- map2stan(
  alist(
    # likelihood
    pulled_left ~ dbinom(1, p),
    
    # linear models
    logit(p) <- A + (BP + BPC * condition) * prosoc_left,
    A <- a + a_actor[actor] + a_block[block_id],
    BP <- bp + bp_actor[actor] + bp_block[block_id],
    BPC <- bpc + bpc_actor[actor] + bpc_block[block_id],
    
    # adaptive priors
    c(a_actor, bp_actor, bpc_actor)[actor] ~ dmvnorm2(0, sigma_actor, Rho_actor),
    c(a_block, bp_block, bpc_block)[block_id] ~ dmvnorm2(0, sigma_block, Rho_block),
    
    # fixed priors
    c(a, bp, bpc) ~ dnorm(0, 1),
    sigma_actor ~ dcauchy(0, 2),
    sigma_block ~ dcauchy(0, 2),
    Rho_actor ~ dlkjcorr(4),
    Rho_block ~ dlkjcorr(4)
  ),
  data = d,
  iter = 5000,
  warmup = 1000,
  chains = 4,
  cores = 4
)

# non-centered parameterization
m13.6NC <- map2stan(
  alist(
    # likelihood
    pulled_left ~ dbinom(1, p),
    
    # linear models
    logit(p) <- A + (BP + BPC * condition) * prosoc_left,
    A <- a + a_actor[actor] + a_block[block_id],
    BP <- bp + bp_actor[actor] + bp_block[block_id],
    BPC <- bpc + bpc_actor[actor] + bpc_block[block_id],
    
    # adaptive NON-CENTERED priors
    c(a_actor, bp_actor, bpc_actor)[actor] ~ dmvnormNC(sigma_actor, Rho_actor),
    c(a_block, bp_block, bpc_block)[block_id] ~ dmvnormNC(sigma_block, Rho_block),
    
    # fixed priors
    c(a, bp, bpc) ~ dnorm(0, 1),
    sigma_actor ~ dcauchy(0, 2),
    sigma_block ~ dcauchy(0, 2),
    Rho_actor ~ dlkjcorr(4),
    Rho_block ~ dlkjcorr(4)
  ),
  data = d,
  iter = 5000,
  warmup = 1000,
  chains = 4,
  cores = 4
)
plot(precis(m13.6NC, depth = 2, pars = c("a_actor", "a_block", "bp_actor", "bp_block")))
```

# Continuous categories and the Gaussian process


